---
title: "Project_2.2_Histone_Modifications_BG_"
author: "Florence Marti"
date: "2025-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence/Histone Modifications Memory")
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
  library(org.Mm.eg.db)
  library(AnnotationDbi)
  library(valr)
  library(EnrichedHeatmap)
  library(gridExtra)
  library(grid)
  library(limma)
  library(plyranges)
  library(dplyr)
})

# Setting my heatmap colour right in the beginning
my_colors <- colorRampPalette(c("blue", "white", "red"))(100)
my_breaks <- seq(-2, 2, length.out = 101)
```

## Importing the downloaded data

```{r}
head(read.table("peaks/H3K27ac.txt", header = TRUE))
```

```{r}
CTCF_df <- read.table("peaks/CTCF.txt", header = TRUE)
H3K27ac_df <- read.table("peaks/H3K27ac.txt", header = TRUE)
H3K27me3_df <- read.table("peaks/H3K27me3.txt", header = TRUE)
H3K4me3_df <- read.table("peaks/H3K4me3.txt", header = TRUE)
H3K9me3_df <- read.table("peaks/H3K9me3.txt", header = TRUE)

# test to see what happens if we exlude these columns
H3K27ac_df <- H3K27ac_df[, -c(11:14)]
H3K27me3_df <- H3K27me3_df[, -c(11:14)]
H3K4me3_df <- H3K4me3_df[, -c(11:14)]
H3K9me3_df <- H3K9me3_df[, -c(11:14)]

# GRanges
CTCF_gr <- GRanges(seqnames = CTCF_df$Chr, ranges = IRanges(start = CTCF_df$Start, end = CTCF_df$End))
H3K27ac_gr <- GRanges(seqnames = H3K27ac_df$Chr, ranges = IRanges(start = H3K27ac_df$Start, end = H3K27ac_df$End))
H3K27me3_gr <- GRanges(seqnames = H3K27me3_df$Chr, ranges = IRanges(start = H3K27me3_df$Start, end = H3K27me3_df$End))
H3K4me3_gr <- GRanges(seqnames = H3K4me3_df$Chr, ranges = IRanges(start = H3K4me3_df$Start, end = H3K4me3_df$End))
H3K9me3_gr <- GRanges(seqnames = H3K9me3_df$Chr, ranges = IRanges(start = H3K9me3_df$Start, end = H3K9me3_df$End))


peaks <- GRangesList(
  CTCF = CTCF_gr,
  H3K27ac = H3K27ac_gr,
  H3K27me3 = H3K27me3_gr,
  H3K4me3 = H3K4me3_gr,
  H3K9me3 = H3K9me3_gr
)

dir.create("peak_bed_files")
rtracklayer::export(peaks$CTCF, "peak_bed_files/CTCF_peaks.bed")
rtracklayer::export(peaks$H3K27ac, "peak_bed_files/H3K27ac_peaks.bed")
rtracklayer::export(peaks$H3K27me3, "peak_bed_files/H3K27me3_peaks.bed")
rtracklayer::export(peaks$H3K4me3, "peak_bed_files/H3K4me3_peaks.bed")
rtracklayer::export(peaks$H3K9me3, "peak_bed_files/H3K9me3_peaks.bed")

```

```{r, fig.cap="Figure 1. UpSet plot showing the overlap of genomic regions (peaks) identified for each histone modification (H3K4me3, H3K27ac, H3K27me3, H3K9me3) in naive and memory T cells. This visualization highlights the extent of shared and unique chromatin regions across hPMTs."}
regionUpset(peaks)
```
In this upset plot we show that while we have a lot of single hits, we also have 7611 bivalent regions for H3K27me3, and H3K4me3. 


## Quality Control

### PCA

```{r}
H3K27ac_PCA <- H3K27ac_df[,7:18]
rownames(H3K27ac_PCA) <- H3K27ac_df$peakN

H3K4me3_PCA <- H3K4me3_df[,7:18]
rownames(H3K4me3_PCA) <- H3K4me3_df$peakN

H3K27me3_PCA <- H3K27me3_df[,7:18]
rownames(H3K27me3_PCA) <- H3K27me3_df$peakN

H3K9me3_PCA <- H3K9me3_df[,7:18]
rownames(H3K9me3_PCA) <- H3K9me3_df$peakN


```

```{r}
run_pca <- function(signal_matrix, mark_name) {
  pca_input <- t(signal_matrix)
  pca_result <- prcomp(pca_input, scale. = TRUE) # scale is important for z scoring
  
  variance <- summary(pca_result)$importance[2, ]
  pca_df <- data.frame(
    PC1 = pca_result$x[, 1],
    PC2 = pca_result$x[, 2],
    CellType = rep(c("Naive_30", "Memory", "Exhausted"), each = 4)
  )
  
  ggplot(pca_df, aes(PC1, PC2, color = CellType)) +
  geom_point(size = 4,) +
  scale_color_manual(values = c("Naive_30" = "#ef77b4", "Memory" = "#ff7f0e", "Exhausted" = "#4daf4a" )) +
    labs(
      title = paste("PCA of", mark_name, "Signal"),
      x = paste0("PC1 (", round(variance[1]*100, 1), "%)"),
      y = paste0("PC2 (", round(variance[2]*100, 1), "%)")
    ) +
  theme_classic()
}



plot_H3K27ac <- run_pca(H3K27ac_PCA, "H3K27ac")
plot_H3K4me3 <- run_pca(H3K4me3_PCA, "H3K4me3")
plot_H3K27me3 <- run_pca(H3K27me3_PCA, "H3K27me3")
plot_H3K9me3 <- run_pca(H3K9me3_PCA, "H3K9me3")
```

```{r}
library(gridExtra)
grid.arrange(
  plot_H3K27ac, plot_H3K4me3,
  plot_H3K27me3, plot_H3K9me3,
  nrow = 2,
  top = "PCA of hPMTs Across all Cell Types"
)
```
```{r}
H3K27ac_PCA <- H3K27ac_PCA %>% tibble::rownames_to_column("peakID")
H3K4me3_PCA <- H3K4me3_PCA %>% tibble::rownames_to_column("peakID")
H3K27me3_PCA <- H3K27me3_PCA %>% tibble::rownames_to_column("peakID")
H3K9me3_PCA <- H3K9me3_PCA %>% tibble::rownames_to_column("peakID")

combined_signals <- list(
  H3K27ac_PCA,
  H3K4me3_PCA,
  H3K27me3_PCA,
  H3K9me3_PCA
) %>% 
  Reduce(function(x, y) inner_join(x, y, by = "peakID"), .)

rownames(combined_signals) <- combined_signals$peakID
combined_signals$peakID <- NULL

dim(combined_signals)

```
```{r}

pca_input <- t(combined_signals) 
pca_result <- prcomp(pca_input, scale. = TRUE)

annotation_col <- data.frame(
  CellType = rep(rep(c("Naive_30", "Memory", "Exhausted"), each = 4), times = 4),
  HistoneMark = rep(c("H3K4me3", "H3K27ac", "H3K27me3", "H3K9me3"), each = 12)
)

rownames(annotation_col) <- colnames(combined_signals)

pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  CellType = annotation_col$CellType,
  HistoneMark = annotation_col$HistoneMark
)

# Plot with custom aesthetics
library(ggplot2)
ggplot(pca_df, aes(PC1, PC2, color = CellType, shape = HistoneMark)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = c("Naive_30" = "#ef77b4", "Memory" = "#ff7f0e", "Exhausted" = "#4daf4a")) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +  # Different shapes for marks
  theme_classic() +
  labs(
    title = "PCA of hPMTs in T Cells",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)")
  ) +
  guides(color = guide_legend(order = 1), shape = guide_legend(order = 2))

```


# Signal Extraction and Filtering
Signal intensities for naive and memory T cells were extracted, and mean values per group were calculated for each peak. We extracted the signal values from the GRanges files for each histone modification and cell type and then computed log2 fold changes and filtered for regions with significant differences (absolute log2(signal) > 0.58496), focusing on biologically relevant changes in chromatin accessibility.

```{r}
# Naive = cols 7:10, Memory = cols 11:14, Exhausted: 15:18
H3K27ac_signal <- cbind(
  naive = H3K27ac_df[, 7:10],
  memory = H3K27ac_df[, 11:14],
  exh = H3K27ac_df[, 15:18]
)
H3K27me3_signal <- cbind(
  naive = H3K27me3_df[, 7:10],
  memory = H3K27me3_df[, 11:14],
  exh = H3K27me3_df[, 15:18]
)
H3K4me3_signal <- cbind(
  naive = H3K4me3_df[, 7:10],
  memory = H3K4me3_df[, 11:14],
  exh = H3K4me3_df[, 15:18]
)

H3K9me3_signal <- cbind(
  naive = H3K9me3_df[, 7:10],
  memory = H3K9me3_df[, 11:14],
  exh = H3K9me3_df[, 15:18]
)


rownames(H3K27ac_signal) <- paste(H3K27ac_df$peakN)
rownames(H3K27me3_signal) <- paste(H3K27me3_df$peakN)
rownames(H3K4me3_signal)  <- paste(H3K4me3_df$peakN)
rownames(H3K9me3_signal)  <- paste(H3K9me3_df$peakN)
```

## Comparing Naive and Memory Cells

```{r}
pseudocount <- 1e-6
## H3K4me3
mean_naive_H3K4me3 <- rowMeans(H3K4me3_signal[, 1:4])   
mean_memory_H3K4me3 <- rowMeans(H3K4me3_signal[, 5:8])  
mean_exh_H3K4me3 <- rowMeans(H3K4me3_signal[, 9:12])

log2_H3K4me3 <- log2((mean_memory_H3K4me3 + pseudocount) / (mean_naive_H3K4me3 + pseudocount))
sig_rows_H3K4me3 <- which(abs(log2_H3K4me3) > 0.58496)
sig_H3K4me3_signal <- H3K4me3_signal[sig_rows_H3K4me3, 1:8]
scaled_signal_H3K4me3 <- t(scale(t(sig_H3K4me3_signal)))  

## H3K27ac
mean_naive_H3K27ac <- rowMeans(H3K27ac_signal[, 1:4])  
mean_memory_H3K27ac <- rowMeans(H3K27ac_signal[, 5:8])  
mean_exh_H3K27ac <- rowMeans(H3K27ac_signal[, 9:12])  

log2_H3K27ac <- log2((mean_memory_H3K27ac + pseudocount) / (mean_naive_H3K27ac + pseudocount))
sig_rows_H3K27ac <- which(abs(log2_H3K27ac) > 0.58496)
sig_H3K27ac_signal <- H3K27ac_signal[sig_rows_H3K27ac,1:8 ]
scaled_signal_H3K27ac <- t(scale(t(sig_H3K27ac_signal)))  

## H3K27me3
mean_naive_H3K27me3 <- rowMeans(H3K27me3_signal[, 1:4])   
mean_memory_H3K27me3 <- rowMeans(H3K27me3_signal[, 5:8]) 
mean_exh_H3K27me3 <- rowMeans(H3K27me3_signal[, 9:12])  

log2_H3K27me3 <- log2((mean_memory_H3K27me3 + pseudocount) / (mean_naive_H3K27me3 + pseudocount))
sig_rows_H3K27me3 <- which(abs(log2_H3K27me3) > 0.58496)
sig_H3K27me3_signal <- H3K27me3_signal[sig_rows_H3K27me3, 1:8 ]
scaled_signal_H3K27me3 <- t(scale(t(sig_H3K27me3_signal)))  


## H3K9me3
mean_naive_H3K9me3 <- rowMeans(H3K9me3_signal[, 1:4])   
mean_memory_H3K9me3 <- rowMeans(H3K9me3_signal[, 5:8]) 
mean_exh_H3K9me3 <- rowMeans(H3K9me3_signal[, 9:12])  

log2_H3K9me3 <- log2((mean_memory_H3K9me3 + pseudocount) / (mean_naive_H3K9me3 + pseudocount))
sig_rows_H3K9me3 <- which(abs(log2_H3K9me3) > 0.58496)
sig_H3K9me3_signal <- H3K9me3_signal[sig_rows_H3K9me3, 1:8 ]
scaled_signal_H3K9me3 <- t(scale(t(sig_H3K9me3_signal)))  
```

## Comparing naive and exhausted cells:
```{r}
## H3K4me3
log2_H3K4me3_EXN <- log2((mean_exh_H3K4me3 + pseudocount) / (mean_naive_H3K4me3 + pseudocount))
sig_rows_H3K4me3_EXN <- which(abs(log2_H3K4me3_EXN) > 0.58496)
sig_H3K4me3_signal_EXN <- H3K4me3_signal[sig_rows_H3K4me3_EXN,c(1:4, 9:12)]
scaled_signal_H3K4me3_EXN <- t(scale(t(sig_H3K4me3_signal_EXN)))  


## H3K27ac
log2_H3K27ac_EXN <- log2((mean_exh_H3K27ac + pseudocount) / (mean_naive_H3K27ac + pseudocount))
sig_rows_H3K27ac_EXN <- which(abs(log2_H3K27ac_EXN) > 0.58496)
sig_H3K27ac_signal_EXN <- H3K27ac_signal[sig_rows_H3K27ac,c(1:4, 9:12)]
scaled_signal_H3K27ac_EXN <- t(scale(t(sig_H3K27ac_signal_EXN)))  

# H3K27me3
log2_H3K27me3_EXN <- log2((mean_exh_H3K27me3 + pseudocount) / (mean_naive_H3K27me3 + pseudocount))
sig_rows_H3K27me3_EXN <- which(abs(log2_H3K27me3_EXN) > 0.58496)
sig_H3K27me3_signal_EXN <- H3K27me3_signal[sig_rows_H3K27me3_EXN,c(1:4, 9:12)]
scaled_signal_H3K27me3_EXN <- t(scale(t(sig_H3K27me3_signal_EXN)))  

# H3K9me3
log2_H3K9me3_EXN <- log2((mean_exh_H3K9me3 + pseudocount) / (mean_naive_H3K9me3 + pseudocount))
sig_rows_H3K9me3_EXN <- which(abs(log2_H3K9me3_EXN) > 0.58496)
sig_H3K9me3_signal_EXN <- H3K9me3_signal[sig_rows_H3K9me3_EXN,c(1:4, 9:12)]
scaled_signal_H3K9me3_EXN <- t(scale(t(sig_H3K9me3_signal_EXN)))  
```


## Comparing memory and exhausted cells:
```{r}
# H3K4me3
log2_H3K4me3_EXM<- log2((mean_exh_H3K4me3 + pseudocount) / (mean_memory_H3K4me3 + pseudocount))
sig_rows_H3K4me3_EXM<- which(abs(log2_H3K4me3_EXM) > 0.58496)
sig_H3K4me3_signal_EXM<- H3K4me3_signal[sig_rows_H3K4me3_EXM, 5:12 ]
scaled_signal_H3K4me3_EXM<- t(scale(t(sig_H3K4me3_signal_EXM)))  


# H3K27ac
log2_H3K27ac_EXM<- log2((mean_exh_H3K27ac + pseudocount) / (mean_memory_H3K27ac + pseudocount))
sig_rows_H3K27ac_EXM<- which(abs(log2_H3K27ac_EXM) > 0.58496)
sig_H3K27ac_signal_EXM<- H3K27ac_signal[sig_rows_H3K27ac, 5:12]
scaled_signal_H3K27ac_EXM<- t(scale(t(sig_H3K27ac_signal_EXM)))  

# H3K27me3
log2_H3K27me3_EXM<- log2((mean_exh_H3K27me3 + pseudocount) / (mean_memory_H3K27me3 + pseudocount))
sig_rows_H3K27me3_EXM<- which(abs(log2_H3K27me3_EXM) > 0.58496)
sig_H3K27me3_signal_EXM<- H3K27me3_signal[sig_rows_H3K27me3_EXM,5:12 ]
scaled_signal_H3K27me3_EXM<- t(scale(t(sig_H3K27me3_signal_EXM)))  

# H3K9me3
log2_H3K9me3_EXM<- log2((mean_exh_H3K9me3 + pseudocount) / (mean_memory_H3K9me3 + pseudocount))
sig_rows_H3K9me3_EXM<- which(abs(log2_H3K9me3_EXM) > 0.58496)
sig_H3K9me3_signal_EXM <- H3K9me3_signal[sig_rows_H3K9me3_EXM, 5:12]
scaled_signal_H3K9me3_EXM<- t(scale(t(sig_H3K9me3_signal_EXM)))  
```


# Visualisation of heatmaps for memory vs naive
```{r}
# Suppose you have 4 naive and 4 memory samples per histone mark
annotation_col <- data.frame(
  CellType = rep(c("Naive T-Cell", "Memory T-Cell"), each = 4)
)

annotation_colors <- list(
  CellType = c("Naive T-Cell" = "#ef77b4", "Memory T-Cell" = "#ff7f0e"))


rownames(annotation_col) <- colnames(scaled_signal_H3K4me3)  
hmap1 <- pheatmap::pheatmap(scaled_signal_H3K4me3, color = my_colors, breaks = my_breaks,  
                            annotation_col = annotation_col,
                            annotation_colors = annotation_colors, 
                            show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)

rownames(annotation_col) <- colnames(scaled_signal_H3K27ac)  
hmap2 <- pheatmap::pheatmap(scaled_signal_H3K27ac, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)


rownames(annotation_col) <- colnames(scaled_signal_H3K27me3)  
hmap3 <- pheatmap::pheatmap(scaled_signal_H3K27me3, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE,  silent = TRUE)

rownames(annotation_col) <- colnames(scaled_signal_H3K9me3)  
hmap4 <- pheatmap::pheatmap(scaled_signal_H3K9me3, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors, show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)
```


# Visualisation of heatmaps for naive vs exhausted
```{r}
# Suppose you have 4 naive and 4 exhausted samples per histone mark
annotation_col <- data.frame(
  CellType = rep(c("Naive T-Cell", "Exhausted T-Cell"), each = 4)
)

annotation_colors <- list(
  CellType = c("Naive T-Cell" = "#ef77b4", "Exhausted T-Cell" = "#4daf4a" ))


rownames(annotation_col) <- colnames(scaled_signal_H3K4me3_EXN)  
hmap5 <- pheatmap::pheatmap(scaled_signal_H3K4me3_EXN, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors, show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)


rownames(annotation_col) <- colnames(scaled_signal_H3K27ac_EXN)  
hmap6 <- pheatmap::pheatmap(scaled_signal_H3K27ac_EXN, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)


rownames(annotation_col) <- colnames(scaled_signal_H3K27me3_EXN)  
hmap7 <- pheatmap::pheatmap(scaled_signal_H3K27me3_EXN, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE,  silent = TRUE)

rownames(annotation_col) <- colnames(scaled_signal_H3K9me3_EXN)  
hmap8 <- pheatmap::pheatmap(scaled_signal_H3K9me3_EXN, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors, show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)
```



# Visualisation of heatmaps for memory vs exhausted
```{r}
# Suppose you have 4 naive and 4 memory samples per histone mark
annotation_col <- data.frame(
  CellType = rep(c("Memory T-Cell", "Exhausted T-Cell"), each = 4)
)

annotation_colors <- list(
  CellType = c("Memory T-Cell" = "#ff7f0e", "Exhausted T-Cell" = "#4daf4a" ))

rownames(annotation_col) <- colnames(scaled_signal_H3K4me3_EXM)  
hmap9 <- pheatmap::pheatmap(scaled_signal_H3K4me3_EXM, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors, show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)


rownames(annotation_col) <- colnames(scaled_signal_H3K27ac_EXM)  
hmap10 <- pheatmap::pheatmap(scaled_signal_H3K27ac_EXM, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)


rownames(annotation_col) <- colnames(scaled_signal_H3K27me3_EXM)  
hmap11 <- pheatmap::pheatmap(scaled_signal_H3K27me3_EXM, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors,show_colnames = FALSE, show_rownames = FALSE,  silent = TRUE)

rownames(annotation_col) <- colnames(scaled_signal_H3K9me3_EXM)  
hmap12 <- pheatmap::pheatmap(scaled_signal_H3K9me3_EXM, color = my_colors, breaks = my_breaks,  annotation_col = annotation_col,annotation_colors = annotation_colors, show_colnames = FALSE, show_rownames = FALSE, silent = TRUE)
```


```{r fig.align='center', fig.height=10, fig.width=12.5, fig.cap="Figure 2. Heatmaps of z-scored signal intensities for significantly different peaks in each histone modification. Each heatmap displays naive, memory, and exhausted T cell samples. Color scale represents scaled signal intensity (blue: low, red: high)"}
# Create histone mark labels as a vertical column
label_grobs <- arrangeGrob(
  textGrob("H3K27ac", rot = 90, gp = gpar(fontsize = 12, fontface = "bold")),
  textGrob("H3K4me3", rot = 90, gp = gpar(fontsize = 12, fontface = "bold")),
  textGrob("H3K27me3", rot = 90, gp = gpar(fontsize = 12, fontface = "bold")),
  textGrob("H3K9me3", rot = 90, gp = gpar(fontsize = 12, fontface = "bold")),
  ncol = 1,
  heights = unit(rep(1, 4), "null")  # match heatmap row heights
)

# Stack the heatmaps as before
col1 <- arrangeGrob(hmap2$gtable, hmap1$gtable, hmap3$gtable, hmap4$gtable, ncol = 1)
col2 <- arrangeGrob(hmap6$gtable, hmap5$gtable, hmap7$gtable, hmap8$gtable, ncol = 1)
col3 <- arrangeGrob(hmap10$gtable, hmap9$gtable, hmap11$gtable, hmap12$gtable, ncol = 1)

# Add titles to each column
col1_with_title <- arrangeGrob(textGrob("Naive vs Memory", gp = gpar(fontsize = 12, fontface = "bold")),
                               col1,
                               ncol = 1,
                               heights = unit.c(unit(1, "lines"), unit(1, "null")))

col2_with_title <- arrangeGrob(textGrob("Naive vs Exhausted", gp = gpar(fontsize = 12, fontface = "bold")),
                               col2,
                               ncol = 1,
                               heights = unit.c(unit(1, "lines"), unit(1, "null")))

col3_with_title <- arrangeGrob(textGrob("Memory vs Exhausted", gp = gpar(fontsize = 12, fontface = "bold")),
                               col3,
                               ncol = 1,
                               heights = unit.c(unit(1, "lines"), unit(1, "null")))

# Combine label column and heatmap columns into full layout
grid.arrange(label_grobs, col1_with_title, col2_with_title, col3_with_title,
             ncol = 4,
             widths = unit.c(unit(1.5, "cm"), unit(1, "null"), unit(1, "null"), unit(1, "null")),
             top = "Histone Modification Profiles Across T Cell States")

```

```{r, eval = T, include = F}
#rm(mean_memory_H3K27ac, mean_naive_H3K27ac, mean_memory_H3K27me3, mean_memory_H3K4me3, mean_memory_H3K9me3, mean_naive_H3K27me3, mean_naive_H3K4me3, mean_naive_H3K9me3, sig_rows_H3K27ac, sig_rows_H3K27me3, sig_rows_H3K4me3, sig_rows_H3K9me3, sig_H3K27ac_signal, sig_H3K27me3_signal,sig_H3K4me3_signal, sig_H3K9me3_signal, log2_H3K27ac, log2_H3K27me3, log2_H3K4me3, log2_H3K9me3, sig_H3K9me3_signal_EXM, sig_rows_H3K9me3_EXM, sig_H3K27me3_signal_EXM, sig_rows_H3K27me3_EXM, sig_H3K27ac_signal_EXM, sig_rows_H3K27ac_EXM, sig_H3K4me3_signal_EXM, sig_rows_H3K4me3_EXM, sig_H3K9me3_signal, sig_rows_H3K9me3, sig_H3K27me3_signal, sig_rows_H3K27me3, sig_H3K27ac_signal, sig_rows_H3K27ac, sig_rows_H3K4me3, sig_H3K4me3_signal)
```

# Comparing all significant hPMTs of different cell type comparisons

# Memory vs Naive

```{r}
# List of all scaled matrices
scaled_list <- list(scaled_signal_H3K4me3, scaled_signal_H3K27ac, scaled_signal_H3K27me3, scaled_signal_H3K9me3)

#rm(scaled_signal_H3K4me3, scaled_signal_H3K27ac, scaled_signal_H3K27me3, scaled_signal_H3K9me3)

# Get union of all rownames (regions)
all_regions <- Reduce(union, lapply(scaled_list, rownames))

# Function to add missing rows with zeros and reorder rows by all_regions
fill_missing_rows <- function(mat, all_rows) {
  missing <- setdiff(all_rows, rownames(mat))
  if(length(missing) > 0) {
    # Create zero matrix for missing rows
    zero_mat <- matrix(0, nrow = length(missing), ncol = ncol(mat),
                       dimnames = list(missing, colnames(mat)))
    # Bind existing matrix and zero rows
    mat <- rbind(mat, zero_mat)
  }
  # Reorder rows
  mat[all_rows, , drop = FALSE]
}

# Apply function to each scaled matrix
scaled_filled <- lapply(scaled_list, fill_missing_rows, all_rows = all_regions)

# Combine all columns together (all samples from all histone mods)
combined_scaled <- do.call(cbind, scaled_filled)

#rm(scaled_list)
```

```{r}
# Get colnames for each scaled matrix (samples)
samples_H3K4me3  <- colnames(scaled_filled[[1]])
samples_H3K27ac  <- colnames(scaled_filled[[2]])
samples_H3K27me3 <- colnames(scaled_filled[[3]])
samples_H3K9me3  <- colnames(scaled_filled[[4]])

all_samples <- c(samples_H3K27ac, samples_H3K4me3, samples_H3K27me3, samples_H3K9me3)


# Assuming 4 naive + 4 memory per histone modification, adjust if different
annotation_col <- data.frame(
  CellType = rep(c("Naive T-Cell", "Memory T-Cell"), times = c(4,4)),
  HistoneMark = rep(c( "H3K27ac", "H3K4me3", "H3K27me3", "H3K9me3"), each = 8)
)

annotation_colors <- list(
  CellType = c("Naive T-Cell" = "#ef77b4", "Memory T-Cell" = "#ff7f0e"),
  HistoneMark = c(
  "H3K4me3"   = "#ffd92f",  
  "H3K27ac"   = "#D73027",  
  "H3K27me3"  = "#a6cee3",  
  "H3K9me3"   = "#6A5ACD"   
))


rownames(annotation_col) <- colnames(combined_scaled)

top_rows <- head(order(apply(combined_scaled, 1, var), decreasing = TRUE), 5000)
```


```{r}
all_hmap1 <- pheatmap::pheatmap(combined_scaled[top_rows, ],
         color = my_colors, breaks = my_breaks, 
         annotation_col = annotation_col,
         annotation_colors = annotation_colors,
         show_colnames = FALSE,
         show_rownames = FALSE,
         main = "Naive vs. Memory T Cell",
         cluster_rows = TRUE,
         cluster_cols = F,
         silent = TRUE)

#rm(top_rows, scaled_filled)
```


# Exhausted vs Naive 

```{r}
# List of all scaled matrices
scaled_list_EXN <- list(scaled_signal_H3K4me3_EXN, scaled_signal_H3K27ac_EXN, scaled_signal_H3K27me3_EXN, scaled_signal_H3K9me3_EXN)

#rm(scaled_signal_H3K4me3_EXN, scaled_signal_H3K27ac_EXN, scaled_signal_H3K27me3_EXN, scaled_signal_H3K9me3_EXN)

# Get union of all rownames (regions)
all_regions_EXN <- Reduce(union, lapply(scaled_list_EXN, rownames))

# Function to add missing rows with zeros and reorder rows by all_regions
fill_missing_rows <- function(mat, all_rows) {
  missing <- setdiff(all_rows, rownames(mat))
  if(length(missing) > 0) {
    # Create zero matrix for missing rows
    zero_mat <- matrix(0, nrow = length(missing), ncol = ncol(mat),
                       dimnames = list(missing, colnames(mat)))
    # Bind existing matrix and zero rows
    mat <- rbind(mat, zero_mat)
  }
  # Reorder rows
  mat[all_rows, , drop = FALSE]
}

# Apply function to each scaled matrix
scaled_filled_EXN <- lapply(scaled_list_EXN, fill_missing_rows, all_rows = all_regions_EXN)

# Combine all columns together (all samples from all histone mods)
combined_scaled_EXN <- do.call(cbind, scaled_filled_EXN)

```

```{r}
# Get colnames for each scaled matrix (samples)
samples_H3K4me3  <- colnames(scaled_filled_EXN[[1]])
samples_H3K27ac  <- colnames(scaled_filled_EXN[[2]])
samples_H3K27me3 <- colnames(scaled_filled_EXN[[3]])
samples_H3K9me3  <- colnames(scaled_filled_EXN[[4]])

all_samples_EXN <- c(samples_H3K27ac, samples_H3K4me3, samples_H3K27me3, samples_H3K9me3)


# Assuming 4 naive + 4 memory per histone modification, adjust if different
annotation_col_2 <- data.frame(
  CellType = rep(c("Naive T-Cell", "Exhausted T-Cell"), times = c(4,4)),
  HistoneMark = rep(c( "H3K27ac", "H3K4me3", "H3K27me3", "H3K9me3"), each = 8)
)

annotation_colors_2 <- list(
  CellType = c("Naive T-Cell" = "#ef77b4", "Exhausted T-Cell" = "#4daf4a"),
  HistoneMark = c(
  "H3K4me3"   = "#ffd92f",  
  "H3K27ac"   = "#D73027",  
  "H3K27me3"  = "#a6cee3",  
  "H3K9me3"   = "#6A5ACD"   
)
)
# 
# rownames(annotation_col_2) <- c("H3K4me3 Naive T-Cell 1 " ,  " H3K4me3 Naive T-Cell 2"  , " H3K4me3 Naive T-Cell 3","H3K4me3 Naive T-Cell 4" ,  " H3K4me3 Exhausted T-Cell 1" , "H3K4me3 Exhausted T-Cell 2","H3K4me3 Exhausted T-Cell 3 ","H3K4me3 Exhausted T-Cell 4","H3K27ac Naive T-Cell 1","H3K27ac Naive T-Cell 2","H3K27ac Naive T-Cell 3","H3K27ac Naive T-Cell 4","H3K27ac Exhausted T-Cell 1","H3K27ac Exhausted T-Cell 2","H3K27ac Exhausted T-Cell 3 ","H3K27ac Exhausted T-Cell 4","H3K27me3 Naive T-Cell 1","H3K27me3 Naive T-Cell 2","H3K27me3 Naive T-Cell 3","H3K27me3 Naive T-Cell 4","H3K27me3 Exhausted T-Cell 1","H3K27me3 Exhausted T-Cell 2","H3K27me3 Exhausted T-Cell 3 ","H3K27me3 Exhausted T-Cell 4","H3K9me3 Naive T-Cell 1","H3K9me3 Naive T-Cell 2","H3K9me3 Naive T-Cell 3","H3K9me3 Naive T-Cell 4","H3K9me3 Exhausted T-Cell 1","H3K9me3 Exhausted T-Cell 2","H3K9me3 Exhausted T-Cell 3 ","H3K9me3 Exhausted T-Cell 4")


rownames(annotation_col_2) <- colnames(combined_scaled_EXN)


top_rows_EXN <- head(order(apply(combined_scaled_EXN, 1, var), decreasing = TRUE), 5000)

all_hmap2 <- pheatmap::pheatmap(combined_scaled_EXN[top_rows_EXN, ],
         color = my_colors, breaks = my_breaks, 
         annotation_col = annotation_col_2,
         annotation_colors = annotation_colors_2,
         show_colnames = FALSE,
         show_rownames = FALSE,
         main = "Exhausted vs. Naive T-Cell",
         cluster_rows = TRUE,
         cluster_cols = F,
         silent = TRUE)

#rm(top_rows, scaled_filled_EXN)
```


```{r}
# List of all scaled matrices
scaled_list_EXM <- list(scaled_signal_H3K4me3_EXM, scaled_signal_H3K27ac_EXM, scaled_signal_H3K27me3_EXM, scaled_signal_H3K9me3_EXM)

#rm(scaled_signal_H3K4me3_EXM, scaled_signal_H3K27ac_EXM, scaled_signal_H3K27me3_EXM, scaled_signal_H3K9me3_EXM)

# Get union of all rownames (regions)
all_regions_EXM <- Reduce(union, lapply(scaled_list_EXM, rownames))

# Function to add missing rows with zeros and reorder rows by all_regions
fill_missing_rows <- function(mat, all_rows) {
  missing <- setdiff(all_rows, rownames(mat))
  if(length(missing) > 0) {
    # Create zero matrix for missing rows
    zero_mat <- matrix(0, nrow = length(missing), ncol = ncol(mat),
                       dimnames = list(missing, colnames(mat)))
    # Bind existing matrix and zero rows
    mat <- rbind(mat, zero_mat)
  }
  # Reorder rows
  mat[all_rows, , drop = FALSE]
}

# Apply function to each scaled matrix
scaled_filled_EXM <- lapply(scaled_list_EXM, fill_missing_rows, all_rows = all_regions_EXM)

# Combine all columns together (all samples from all histone mods)
combined_scaled_EXM <- do.call(cbind, scaled_filled_EXM)
```


```{r}
# Get colnames for each scaled matrix (samples)
samples_H3K4me3  <- colnames(scaled_filled_EXM[[1]])
samples_H3K27ac  <- colnames(scaled_filled_EXM[[2]])
samples_H3K27me3 <- colnames(scaled_filled_EXM[[3]])
samples_H3K9me3  <- colnames(scaled_filled_EXM[[4]])

all_samples_EXM <- c(samples_H3K27ac, samples_H3K4me3, samples_H3K27me3, samples_H3K9me3)


# Assuming 4 naive + 4 memory per histone modification, adjust if different
annotation_col_3 <- data.frame(
  CellType = rep(c("Memory T-Cell", "Exhausted T-Cell"), times = c(4,4)),
  HistoneMark = rep(c( "H3K27ac", "H3K4me3", "H3K27me3", "H3K9me3"), each = 8))

annotation_colors_3 <- list(
  CellType = c("Memory T-Cell" = "#ff7f0e","Exhausted T-Cell"= "#4daf4a"),
  HistoneMark = c(
  "H3K4me3"   = "#ffd92f",  
  "H3K27ac"   = "#D73027",  
  "H3K27me3"  = "#a6cee3",  
  "H3K9me3"   = "#6A5ACD"   
)
)


rownames(annotation_col_3) <- colnames(combined_scaled_EXM)

top_rows_EXM <- head(order(apply(combined_scaled_EXM, 1, var), decreasing = TRUE), 5000)


all_hmap3 <- pheatmap::pheatmap(combined_scaled_EXM[top_rows_EXM, ],
         color = my_colors, breaks = my_breaks, 
         annotation_col = annotation_col_3,
         annotation_colors = annotation_colors_3,
         show_colnames = FALSE,
         show_rownames = FALSE,
         main = "Memory vs. Exhausted T-Cell",
         cluster_rows = TRUE,
         cluster_cols = F,
         silent = TRUE)

#rm(top_rows, scaled_filled_EXM)
```


```{r, fig.width=10, fig.height=20}
grid.arrange(
  all_hmap1$gtable, 
  all_hmap2$gtable, 
  all_hmap3$gtable, 
  ncol = 1,
  top = "Scaled Signal Intensity Across hPMTs"
)
```


# Filtering and Exporting Regions of Interest for potential later use 

### Naive and Memory 

```{r}

H3K4me3_MN  <- combined_scaled[, 5:8]
H3K27ac_MN  <- combined_scaled[, 13:16]
H3K27me3_MN <- combined_scaled[, 21:24]
H3K9me3_MN  <- combined_scaled[, 29:32]

# Activating High
open_high_H3K4me3_MN  <- apply(H3K4me3_MN, 1, function(x) all(x >= 0))
open_high_H3K27ac_MN  <- apply(H3K27ac_MN, 1, function(x) all(x >= 0))

select_rows_MN <- which(open_high_H3K4me3_MN & open_high_H3K27ac_MN)

memory_activating_MN <- combined_scaled[select_rows_MN, c(5:8, 13:16, 21:24, 29:32)]

memory_activating_MN <- memory_activating_MN[!apply(memory_activating_MN == 0, 1, any), ]
dim(memory_activating_MN)

write.csv(memory_activating_MN, file = "files/memory_open_peaks_signal.csv", row.names = T)
```

```{r}
# Repressed High
closed_high_H3K27me3_MN <- apply(H3K27me3_MN, 1, function(x) all(x >= 0))
closed_high_H3K9me3_MN  <- apply(H3K9me3_MN, 1, function(x) all(x >= 0))

select_rows_rep_MN <- which(closed_high_H3K27me3_MN & closed_high_H3K9me3_MN)

memory_repressing_MN <- combined_scaled[select_rows_rep_MN, c(5:8, 13:16, 21:24, 29:32)]
memory_repressing_MN <- memory_repressing_MN[!apply(memory_repressing_MN == 0, 1, any), ]

dim(memory_repressing_MN)

write.csv(memory_repressing_MN, file= "files/memory_closed_peaks_signal.csv", row.names = T)
```

```{r}
# Bivalent 

select_rows_b_MN <- which(open_high_H3K4me3_MN & closed_high_H3K27me3_MN)

memory_bivalent_MN <- combined_scaled[select_rows_b_MN, c(5:8, 13:16, 21:24, 29:32)]
memory_bivalent_MN <- memory_bivalent_MN[!apply(memory_bivalent_MN == 0, 1, any), ]

dim(memory_bivalent_MN)

write.csv(memory_bivalent_MN, file = "files/memory_bivalent.csv", row.names = T)
```
Using this approach we decrease the size of the dataset for comparing naive and memory T cells to 308 peaks, which is substantially less than the initial 45174 peaks we looked at.

### Naive and Exhausted
```{r}
# Activating High
H3K4me3_EXN  <- combined_scaled_EXN[, 5:8]
H3K27ac_EXN  <- combined_scaled_EXN[, 13:16]
H3K27me3_EXN <- combined_scaled_EXN[, 21:24]
H3K9me3_EXN  <- combined_scaled_EXN[, 29:32]

open_high_H3K4me3_EXN  <- apply(H3K4me3_EXN, 1, function(x) all(x >= 0))
open_high_H3K27ac_EXN  <- apply(H3K27ac_EXN, 1, function(x) all(x >= 0))

select_rows_EXN <- which(open_high_H3K4me3_EXN & open_high_H3K27ac_EXN) 
activating_EXN <- combined_scaled_EXN[select_rows_EXN, c(5:8, 13:16, 21:24, 29:32)]
activating_EXN <- activating_EXN[!apply(activating_EXN == 0, 1, any), ]

dim(activating_EXN)

write.csv(activating_EXN, file = "files/open_peaks_signal_EXN.csv", row.names = T)
```

```{r}
# Repressing High
closed_high_H3K27me3_EXN <- apply(H3K27me3_EXN, 1, function(x) all(x >= 0))
closed_high_H3K9me3_EXN  <- apply(H3K9me3_EXN, 1, function(x) all(x >= 0))


select_rows_rep_EXN <- which(closed_high_H3K27me3_EXN & closed_high_H3K9me3_EXN) 

repressing_EXN <- combined_scaled_EXN[select_rows_rep_EXN, c(5:8, 13:16, 21:24, 29:32)]
repressing_EXN <- repressing_EXN[!apply(repressing_EXN == 0, 1, any), ]

dim(repressing_EXN)

write.csv(repressing_EXN, file= "files/closed_peaks_signal_EXN.csv", row.names = T)
```


```{r}
# Bivalent
select_rows_b_EXN <- which(open_high_H3K4me3_EXN & closed_high_H3K27me3_EXN)
bivalent_EXN <- combined_scaled_EXN[select_rows_b_EXN, c(5:8, 13:16, 21:24, 29:32)]
bivalent_EXN <- bivalent_EXN[!apply(bivalent_EXN == 0, 1, any), ]

dim(bivalent_EXN)

write.csv(bivalent_EXN, file = "files/bivalent_EXN.csv", row.names = T)
```


### Exhausted and Memory
```{r}
# Activating High
H3K4me3_EXM  <- combined_scaled_EXM[, 5:8]
H3K27ac_EXM  <- combined_scaled_EXM[, 13:16]
H3K27me3_EXM <- combined_scaled_EXM[, 21:24]
H3K9me3_EXM  <- combined_scaled_EXM[, 29:32]

open_high_H3K4me3_EXM  <- apply(H3K4me3_EXM, 1, function(x) all(x >= 0))
open_high_H3K27ac_EXM  <- apply(H3K27ac_EXM, 1, function(x) all(x >= 0))

select_rows_EXM <- which(open_high_H3K4me3_EXM & open_high_H3K27ac_EXM) 

activating_EXM <- combined_scaled_EXM[select_rows_EXM, c(5:8, 13:16, 21:24, 29:32)]
activating_EXM <- activating_EXM[!apply(activating_EXM == 0, 1, any), ]

dim(activating_EXM)

write.csv(activating_EXM, file = "files/open_peaks_signal_EXM.csv", row.names = T)
```

```{r}
# Repressing High

closed_high_H3K27me3_EXM <- apply(H3K27me3_EXM, 1, function(x) all(x >= 0))
closed_high_H3K9me3_EXM  <- apply(H3K9me3_EXM, 1, function(x) all(x >= 0))

select_rows_rep_EXM <- which(closed_high_H3K27me3_EXM & closed_high_H3K9me3_EXM) 

repressing_EXM <- combined_scaled[select_rows_rep_EXM, c(5:8, 13:16, 21:24, 29:32)]

repressing_EXM <- repressing_EXM[!apply(repressing_EXM == 0, 1, any), ]

dim(repressing_EXM)
write.csv(repressing_EXM, file= "files/closed_peaks_signal_EXM.csv", row.names = T)
```


```{r}
# Bivalent
select_rows_b_EXM <- which(open_high_H3K4me3_EXM & closed_high_H3K27me3_EXM)
bivalent_EXM <- combined_scaled[select_rows_b_EXM, c(5:8, 13:16, 21:24, 29:32)]
bivalent_EXM <- bivalent_EXM[!apply(bivalent_EXM == 0, 1, any), ]

dim(bivalent_EXM)

write.csv(bivalent_EXM, file = "files/bivalent_EXM.csv", row.names = T)
```


```{r, warning=T, include = F, eval=F}
#rm(list = ls())
```
