---
title: "BW_Files_Analysis_Project"
author: "Florence Marti"
date: "2025-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence")
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
  library(org.Mm.eg.db)
  library(AnnotationDbi)
  library(valr)
  library(EnrichedHeatmap)
  library(gridExtra)
  library(grid)
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(PWMEnrich)
})
```

```{r}
ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
genes <- genes(ensdb)
```

# Comparing the histone modifications of naive T cells
First reinporting the GRanges files of all the peaks
```{r}
peaks_part1 <- GRangesList(
  H3K27ac = import("peak_bed_files/H3K27ac_peaks.bed"),
  H3K27me3 = import("peak_bed_files/H3K27me3_peaks.bed"),
  H3K4me3 = import("peak_bed_files/H3K4me3_peaks.bed"),
  H3K9me3 = import("peak_bed_files/H3K9me3_peaks.bed")
)

```
When trying to apply the filters from the lecture (w10), we were left with 0 values. this is why we proceeded with all of the values 
- one idea might be to apply the significant rows that we found in the part prior

```{r}
# Naive
bwfiles_naive <- list.files("bigwig_files", pattern = "_naive.bw$", full.names = TRUE)
names(bwfiles_naive) <- gsub("\\.bw$", "", basename(bwfiles_naive))
naive_regions <- reduce(unlist(GRangesList(peaks_part1)))
sm_n <- signal2Matrix(bwfiles_naive, regions = naive_regions, w = 40)

# Memory
bwfiles_memory <- list.files("bigwig_files", pattern = "_memory.bw$", full.names = TRUE)
names(bwfiles_memory) <- gsub("\\.bw$", "", basename(bwfiles_memory))
memory_regions <- reduce(unlist(GRangesList(peaks_part1)))
sm_m <- signal2Matrix(bwfiles_memory, regions = memory_regions, w = 40)

# Exhausted
bwfiles_exh <- list.files("bigwig_files", pattern = "_EX.bw$", full.names = TRUE)
names(bwfiles_exh) <- gsub("\\.bw$", "", basename(bwfiles_exh))
exh_regions <- reduce(unlist(GRangesList(peaks_part1)))
sm_e <- signal2Matrix(bwfiles_exh, regions = exh_regions, w = 40)

```

```{r}
sm_m1 <- sm_m[1:5000,]
sm_n1 <- sm_n[1:5000,]
sm_e1<- sm_e[1:5000,]
```


```{r, fig.height= 5, fig.width=10}
plotEnrichedHeatmaps(sm_n1, multiScale = T, cluster_rows = T, show_row_dend = T)
```


```{r, fig.height= 5, fig.width=10}
plotEnrichedHeatmaps(sm_m1,  multiScale = T, cluster_rows = T, show_row_dend = T)
```


```{r, fig.height= 5, fig.width=10}
plotEnrichedHeatmaps(sm_e1,  multiScale = T, cluster_rows = T, show_row_dend = T)
```

# Clustering

Finding the best kmeans for clustering of the samples
```{r}
cl_n <- clusterSignalMatrices(sm_n1, k=2:7)
cluster1 <- ggplot(cl_n$varExplained, aes(k, varExplained)) + geom_line() + labs(x = "Number of Clusters (k)", y = "Variance Explained") + ggtitle("Cluster Signal Matrix Naive T-Cells")  + theme_minimal()

cl_m <- clusterSignalMatrices(sm_m1, k=2:7)
cluster2 <- ggplot(cl_m$varExplained, aes(k, varExplained)) + geom_line() + labs(x = "Number of Clusters (k)", y = "Variance Explained") + ggtitle("Cluster Signal Matrix Memory T-Cells") + theme_minimal()

cl_e <- clusterSignalMatrices(sm_e1, k=2:7)
cluster3 <- ggplot(cl_e$varExplained, aes(k, varExplained)) + geom_line() + labs(x = "Number of Clusters (k)", y = "Variance Explained")+ ggtitle("Cluster Signal Matrix Exhausted T-Cells") + theme_minimal()
```
```{r, fig.height= 10, fig.width=10}
grid.arrange(cluster1, cluster2, cluster3,
             ncol = 1,
             top = "K-Means Clustering: Variance Explained per k")

```



```{r}
set.seed(123)  # to ensure that it gives the same results everytime
cl_k_n <- clusterSignalMatrices(sm_n1, k=6)
cl_k_m <- clusterSignalMatrices(sm_m1, k=6)
cl_k_e <- clusterSignalMatrices(sm_e1, k=6)
```
using k = 6 we have an average 95% of the variance explained.

## Adding clusters to the signal matrices and plotting the enriched heatmaps

```{r}
rowData(sm_n1)$cluster <- cl_k_n
rowData(sm_m1)$cluster <- cl_k_m
rowData(sm_e1)$cluster <- cl_k_e

mycolors <- c("1"="pink", "2"="lightblue", "3"="darkgreen", "4"="orange", "5"="purple", "6"="blue")
plotEnrichedHeatmaps(sm_n1, row_split="cluster", mean_color=mycolors, colors=c("white","darkred"))
```


```{r}
plotEnrichedHeatmaps(sm_m1, row_split="cluster", mean_color=mycolors, colors=c("white","darkred"))
```


```{r}
plotEnrichedHeatmaps(sm_e1, row_split="cluster", mean_color=mycolors, colors=c("white","darkred"))
```

## 

```{r}
naive <- meltSignals(sm_n1, splitBy=cl_k_n)
memory <- meltSignals(sm_m1, splitBy=cl_k_m)
exhausted <- meltSignals(sm_e1, splitBy=cl_k_e)

melt1 <- ggplot(naive, aes(position, mean, colour=sample)) + geom_line(size=1.2) + facet_wrap(~split) + ggtitle("Naive T-Cells")
melt2 <- ggplot(memory, aes(position, mean, colour=sample)) + geom_line(size=1.2) + facet_wrap(~split) + ggtitle("Memory T-Cells")
melt3 <- ggplot(exhausted, aes(position, mean, colour=sample)) + geom_line(size=1.2) + facet_wrap(~split) + ggtitle("Exhausted T-Cells")
```

```{r, fig.height = 10, fig.width=8}
library(cowplot)

plot_grid(melt1, melt2, melt3, ncol = 1)
```

