---
title: "Project_Valerie_Florence_EpiGenetics"
author: "Valerie Heckendorn and Florence Marti"
date: "2025-05-07"
output: html_document
---

# Introduction and background

The publication shared the data and compared the naive, memory and exhausted T cells at day 30 and day 32. They also provided 4 replicates of each timepoint, for simplicity reasons, we chose to download rep x for the analysis and then if time allows to reanalyse with another rep.

We downloaded the 4 histone modifications they provided for all 3 cell types, all two days, resulting in 24

# Setup and download
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence")
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
})
```


## Peaks and Histone modifications
```{r}
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence/Histone Modifications Memory")

download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K27ac%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K27ac_wID.txt.gz")
gunzip("peaks/NormRC_H3K27ac_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K27me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K27me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K27me3_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K4me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K4me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K4me3_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K9me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K9me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K9me3_wID.txt.gz")

```

The peaks and histone modifications were not published under the same accession number.

For Naive and Memory cell comparison, day 30 and replicate 1 were chosen. 

### Naive T cells
```{r}
options(timeout = 12000)

# Key Activating

#H3K4me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699095&format=file&file=GSM8699095%5FNaive%5FP14%5FDay30%5FH3K4me3%5FRep1%5FS1%2Escale%2Ebw", "H3K4me3_naive_30.bw.gz")
gunzip("H3K4me3_naive_30.bw.gz")


#H3K27ac
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699079&format=file&file=GSM8699079%5FNaive%5FP14%5FDay30%5FH3K27ac%5FRep1%5FS1%2Escale%2Ebw", "H3K27ac_naive_30.bw.gz")
gunzip("H3K27ac_naive_30.bw.gz")

# KeyRepressive

#H3K27me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699111&format=file&file=GSM8699111%5FNaive%5FP14%5FDay30%5FH3K27me3%5FRep2%5FS6%2Escale%2Ebw", "H3K27me3_naive_30.bw.gz")
gunzip("H3K27me3_naive_30.bw.gz")

#H3K9me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699127&format=file&file=GSM8699127%5FNaive%5FP14%5FDay30%5FH3K9me3%5FRep1%5FS1%2Escale%2Ebw", "H3K9me3_naive_30.bw.gz")
gunzip("H3K9me3_naive_30.bw.gz")


```
### T mem 

```{r}
options(timeout = 12000)

#H3K4me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699103&format=file&file=GSM8699103%5FArm%5FP14%5FDay30%5FH3K4me3%5FRep1%5FS5%2Escale%2Ebw", "H3K4me3_memory.bw.gz")
gunzip("H3K4me3_memory.bw.gz")


#H3K27ac
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699087&format=file&file=GSM8699087%5FArm%5FP14%5FDay30%5FH3K27ac%5FRep1%5FS5%2Escale%2Ebw", "H3K27ac_memory.bw.gz")
gunzip("H3K27ac_memory.bw.gz")

# KeyRepressive

#H3K27me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699119&format=file&file=GSM8699119%5FArm%5FP14%5FDay30%5FH3K27me3%5FRep1%5FS7%2Escale%2Ebw", "H3K27me3_memory.bw.gz")
gunzip("H3K27me3_memory.bw.gz")

#H3K9me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699135&format=file&file=GSM8699135%5FArm%5FP14%5FDay30%5FH3K9me3%5FRep1%5FS3%2Escale%2Ebw", "H3K9me3_memory.bw.gz")
gunzip("H3K9me3_memory.bw.gz")

```

### T exh

```{r}
options(timeout = 12000)

#H3K27ac, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699091&format=file&file=GSM8699091%5FCl13%5FP14%5FDay32%5FH3K27ac%5FRep1%5FS13%2Escale%2Ebw", "H3K27ac_EX.bw")
H3K27ac_EX <- rtracklayer::import("H3K27ac_EX.bw", format = "bw")

#H3K27me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699123&format=file&file=GSM8699123%5FCl13%5FP14%5FDay32%5FH3K27me3%5FRep1%5FS9%2Escale%2Ebw", "H3K27me3_EX.bw")
H3K27me3_EX <- rtracklayer::import("H3K27me3_EX.bw", format = "bw")

#H3K4me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699107&format=file&file=GSM8699107%5FCl13%5FP14%5FDay32%5FH3K4me3%5FRep1%5FS13%2Escale%2Ebw", "H3K4me3_EX.bw")
H3K4me3_EX <- rtracklayer::import("H3K4me3_EX.bw", format = "bw")

#H3K9me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699139&format=file&file=GSM8699139%5FCl13%5FP14%5FDay32%5FH3K9me3%5FRep1%5FS5%2Escale%2Ebw", "H3K9me3_EX.bw")
H3K9me3_EX <- rtracklayer::import("H3K9me3_EX.bw", format = "bw")
```


## Bibliography
