---
title: "Project_Valerie_Florence_EpiGenetics"
author: "Valerie Heckendorn and Florence Marti"
date: "2025-05-07"
output: 
  html_document:
    citation_package: natbib
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence")
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
})
```


# Introduction and background
Based on a mutual interest in immune cells, and data availability, we decided to partially redo the analysis of the paper "Deciphering the role of Histone Modifications in Memory and Exhausted CD8 T Cells" by Huang et al. The paper was published on May 19th 2025 in Scientific Reports. [@Huang2025]

T Lymphocytes (T Cells) are immune cells found in the blood. They are the cell component of the adaptive immune system and are responsible for mediating cell-based immune responses. Naïve T cells are developed in the thymus and can be grouped into many subgroups, amongst them CD8+ cells, which this analysis focuses on. Upon recognition of antigens by receptors on the cell surface, naïve CD8+ T cells (T$_{naive}$), further referenced naïve T cells, undergo activation, clonal expansion, and differentiation into mature T cells. Once matured, T cells can take up their intended function of killing infected cells, producing cytokines, and regulating immune responses. A small portion of mature T cells differentiates into memory T cells (T$_{Mem}$), upon antigen re-exposure, they show rapid effector function and provide the host long term protection. [@Sun2023]

When mature T cells are exposed to chronic infections and cancer, T cells can enter a state of exhaustion, and are no longer equally functional as before. Exhausted T cells (T$_{Exh}$) differ from mature T Cells or T$_{Mem}$ on a transcriptomic and epigenetic level. [@Wherry2011][@Wherry2010]

Compared to T$_{Mem}$, T$_{Exh}$ show a unique open chromatin landscape, which was explored by comparing histone modification of T$_{naive}$, T$_{Mem}$, and T$_{Exh}$. The four histone modifications that were analysed are H3K3


We downloaded the 4 histone modifications they provided for all 3 cell types, all two days, resulting in 24



## Peaks and Histone modifications
```{r}
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K27ac%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K27ac_wID.txt.gz")
gunzip("peaks/NormRC_H3K27ac_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K27me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K27me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K27me3_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K4me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K4me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K4me3_wID.txt.gz")
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE285245&format=file&file=GSE285245%5FNormRC%5FH3K9me3%5FwID%2Etxt%2Egz", "peaks/NormRC_H3K9me3_wID.txt.gz")
gunzip("peaks/NormRC_H3K9me3_wID.txt.gz")

```

The peaks and histone modifications were not published under the same accession number.

For Naive and Memory cell comparison, day 30 and replicate 1 were chosen. 

### Naive T cells
```{r}
options(timeout = 12000)

# Key Activating

#H3K4me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699095&format=file&file=GSM8699095%5FNaive%5FP14%5FDay30%5FH3K4me3%5FRep1%5FS1%2Escale%2Ebw", "H3K4me3_naive_30.bw.gz")
gunzip("H3K4me3_naive_30.bw.gz")


#H3K27ac
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699079&format=file&file=GSM8699079%5FNaive%5FP14%5FDay30%5FH3K27ac%5FRep1%5FS1%2Escale%2Ebw", "H3K27ac_naive_30.bw.gz")
gunzip("H3K27ac_naive_30.bw.gz")

# KeyRepressive

#H3K27me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699111&format=file&file=GSM8699111%5FNaive%5FP14%5FDay30%5FH3K27me3%5FRep2%5FS6%2Escale%2Ebw", "H3K27me3_naive_30.bw.gz")
gunzip("H3K27me3_naive_30.bw.gz")

#H3K9me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699127&format=file&file=GSM8699127%5FNaive%5FP14%5FDay30%5FH3K9me3%5FRep1%5FS1%2Escale%2Ebw", "H3K9me3_naive_30.bw.gz")
gunzip("H3K9me3_naive_30.bw.gz")


```
### T mem 

```{r}
options(timeout = 12000)

#H3K4me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699103&format=file&file=GSM8699103%5FArm%5FP14%5FDay30%5FH3K4me3%5FRep1%5FS5%2Escale%2Ebw", "H3K4me3_memory.bw.gz")
gunzip("H3K4me3_memory.bw.gz")


#H3K27ac
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699087&format=file&file=GSM8699087%5FArm%5FP14%5FDay30%5FH3K27ac%5FRep1%5FS5%2Escale%2Ebw", "H3K27ac_memory.bw.gz")
gunzip("H3K27ac_memory.bw.gz")

# KeyRepressive

#H3K27me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699119&format=file&file=GSM8699119%5FArm%5FP14%5FDay30%5FH3K27me3%5FRep1%5FS7%2Escale%2Ebw", "H3K27me3_memory.bw.gz")
gunzip("H3K27me3_memory.bw.gz")

#H3K9me3
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699135&format=file&file=GSM8699135%5FArm%5FP14%5FDay30%5FH3K9me3%5FRep1%5FS3%2Escale%2Ebw", "H3K9me3_memory.bw.gz")
gunzip("H3K9me3_memory.bw.gz")

```

### T exh

```{r}
options(timeout = 12000)

#H3K27ac, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699091&format=file&file=GSM8699091%5FCl13%5FP14%5FDay32%5FH3K27ac%5FRep1%5FS13%2Escale%2Ebw", "H3K27ac_EX.bw")
H3K27ac_EX <- rtracklayer::import("H3K27ac_EX.bw", format = "bw")

#H3K27me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699123&format=file&file=GSM8699123%5FCl13%5FP14%5FDay32%5FH3K27me3%5FRep1%5FS9%2Escale%2Ebw", "H3K27me3_EX.bw")
H3K27me3_EX <- rtracklayer::import("H3K27me3_EX.bw", format = "bw")

#H3K4me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699107&format=file&file=GSM8699107%5FCl13%5FP14%5FDay32%5FH3K4me3%5FRep1%5FS13%2Escale%2Ebw", "H3K4me3_EX.bw")
H3K4me3_EX <- rtracklayer::import("H3K4me3_EX.bw", format = "bw")

#H3K9me3, Day32, rep1
download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSM8699139&format=file&file=GSM8699139%5FCl13%5FP14%5FDay32%5FH3K9me3%5FRep1%5FS5%2Escale%2Ebw", "H3K9me3_EX.bw")
H3K9me3_EX <- rtracklayer::import("H3K9me3_EX.bw", format = "bw")

```


## Bibliography
