---
title: "Histone Analysis"
author: "Valerie Heckendorn"
date: "2025-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)¨
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence/Histone_Modification_EXH/peaks")
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(valr)
  library(EnrichedHeatmap)
  library(ChIPseeker)
})
```

# Importing the downloaded data
The histone modifications were shared on GEO Accession Viewer as TXT files.

For the analysis, we split the workload in comparing naive and memory cells (part 1), and naive and exhausted cells (part 2). 



```{r}
head(read.table("peaks/NormRC_H3K27ac_wID.txt", header = TRUE))
```
```{r}
seqlevelsStyle(peaks) <- "UCSC"
annotate_peaks_with_genes <- function(gr, txdb, annoDb) {
  # Perform annotation
  annotated <- annotatePeak(gr, TxDb = txdb, annoDb = annoDb)
  
  # Convert to data frame
  df <- as.data.frame(annotated)
  
  # Map geneId (Entrez) to SYMBOL if missing
  if (!"SYMBOL" %in% colnames(df) && "geneId" %in% colnames(df)) {
    df$SYMBOL <- mapIds(org.Hs.eg.db, 
                        keys = df$geneId,
                        column = "SYMBOL",
                        keytype = "ENTREZID",
                        multiVals = "first")
  }
  
  return(df)
}

# Annotate each GRanges object
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
annoDb <- "org.Hs.eg.db"

H3K27ac_annot <- annotate_peaks_with_genes(peaks$H3K27ac, txdb, annoDb)
H3K27me3_annot <- annotate_peaks_with_genes(peaks$H3K27me3, txdb, annoDb)
H3K4me3_annot  <- annotate_peaks_with_genes(peaks$H3K4me3, txdb, annoDb)
H3K9me3_annot  <- annotate_peaks_with_genes(peaks$H3K9me3, txdb, annoDb)

# Optionally inspect the first few annotated rows
head(H3K27me3_annot[, c("seqnames", "start", "end", "annotation", "geneId", "SYMBOL", "distanceToTSS")])


```



```{r}
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence/Histone_Modification_EXH/peaks")

# H3K27ac
H3K27ac_df <- read.table("peaks/NormRC_H3K27ac_wID.txt", header = TRUE)

# H3K27me3
H3K27me3_df <- read.table("peaks/NormRC_H3K27me3_wID.txt", header = TRUE)

# H3K4me3
H3K4me3_df <- read.table("peaks/NormRC_H3K4me3_wID.txt", header = TRUE)

# H3K9me3
H3K9me3_df <- read.table("peaks/NormRC_H3K9me3_wID.txt", header = TRUE)

# Converting to GRanges

H3K27ac_gr <- GRanges(seqnames = H3K27ac_df$Chr, ranges = IRanges(start = H3K27ac_df$Start, end = H3K27ac_df$End))
H3K27me3_gr <- GRanges(seqnames = H3K27me3_df$Chr, ranges = IRanges(start = H3K27me3_df$Start, end = H3K27me3_df$End))
H3K4me3_gr <- GRanges(seqnames = H3K4me3_df$Chr, ranges = IRanges(start = H3K4me3_df$Start, end = H3K4me3_df$End))
H3K9me3_gr <- GRanges(seqnames = H3K9me3_df$Chr, ranges = IRanges(start = H3K9me3_df$Start, end = H3K9me3_df$End))

# Create GRanges List

peaks <- GRangesList(
  H3K27ac = H3K27ac_gr,
  H3K27me3 = H3K27me3_gr,
  H3K4me3 = H3K4me3_gr,
  H3K9me3 = H3K9me3_gr
)

```

# We want to visualize changes in histone modifications between the different cells (here: part 2, naive vs. exhausted). The intensity is defined as the mean expression of the histones over the 4 reps. We then want to plot only regions with significant changes, which we defined as log2 fold change >0.1 or <-0.1.

#Annotate Peaks with Gene Information
```{r}

```


# H3K27ac
```{r}
H3K27ac_new <- H3K27ac_df[,-c(1:10, 15:18)]

mean_naive_H3K27ac <- rowMeans(H3K27ac_new[, 1:4])   # naive day 32 cols
mean_exh_H3K27ac <- rowMeans(H3K27ac_new[, 5:8])  #  exhausted day 32 cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K27ac <- log2((mean_exh_H3K27ac + pseudocount) / (mean_naive_H3K27ac + pseudocount))
summary(log2FC_H3K27ac)

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K27ac <- which(abs(log2FC_H3K27ac) > 0.1)

sig_H3K27ac_signal <- H3K27ac_new[sig_rows_H3K27ac, ]

# Scale the signal matrix for plotting
scaled_signal_H3K27ac <- t(scale(t(sig_H3K27ac_signal)))  
# Scale rows (regions) - so each region is mean 0, sd 1

pheatmap(scaled_signal_H3K27ac, main = "H3K27ac (Naive & Exhausted) Significant Regions", show_rownames = FALSE, cutree_rows = 2)

```
# H3K27me3
```{r}
H3K27me3_new <- H3K27me3_df[,-c(1:10, 15:18)]

mean_naive_H3K27me3 <- rowMeans(H3K27me3_new[, 1:4])   # naive day 32 cols
mean_exh_H3K27me3 <- rowMeans(H3K27me3_new[, 5:8])  #  exhausted day 32 cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K27me3 <- log2((mean_exh_H3K27me3 + pseudocount) / (mean_naive_H3K27me3 + pseudocount))
summary(log2FC_H3K27me3)

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K27me3 <- which(abs(log2FC_H3K27me3) > 0.1)

sig_H3K27me3_signal <- H3K27me3_new[sig_rows_H3K27me3, ]

# Scale the signal matrix for plotting
scaled_signal_H3K27me3 <- t(scale(t(sig_H3K27me3_signal)))  
# Scale rows (regions) - so each region is mean 0, sd 1

pheatmap(scaled_signal_H3K27me3, main = "H3K27me3 (Naive & Exhausted) Significant Regions", show_rownames = FALSE, cutree_rows = 2)

```

#H3K4me3
```{r}
H3K4me3_new <- H3K4me3_df[,-c(1:10, 15:18)]

mean_naive_H3K4me3 <- rowMeans(H3K4me3_new[, 1:4])   # naive day 32 cols
mean_exh_H3K4me3 <- rowMeans(H3K4me3_new[, 5:8])  #  exhausted day 32 cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K4me3 <- log2((mean_exh_H3K4me3+ pseudocount) / (mean_naive_H3K4me3 + pseudocount))
summary(log2FC_H3K4me3)

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K4me3 <- which(abs(log2FC_H3K4me3) > 0.1)

sig_H3K4me3_signal <- H3K4me3_new[sig_rows_H3K4me3, ]

# Scale the signal matrix for plotting
scaled_signal_H3K4me3 <- t(scale(t(sig_H3K4me3_signal)))  
# Scale rows (regions) - so each region is mean 0, sd 1

pheatmap(scaled_signal_H3K4me3, main = "H3K4me3 (Naive & Exhausted) Significant Regions", show_rownames = FALSE, cutree_rows = 2)

```

#H3K9me3
```{r}
H3K9me3_new <- H3K9me3_df[,-c(1:10, 15:18)]

mean_naive_H3K9me3 <- rowMeans(H3K9me3_new[, 1:4])   # naive day 32 cols
mean_exh_H3K9me3 <- rowMeans(H3K9me3_new[, 5:8])  #  exhausted day 32 cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K9me3 <- log2((mean_exh_H3K9me3 + pseudocount) / (mean_naive_H3K9me3 + pseudocount))
summary(log2FC_H3K9me3)

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K9me3 <- which(abs(log2FC_H3K9me3) > 0.1)

sig_H3K9me3_signal <- H3K9me3_new[sig_rows_H3K9me3, ]

# Scale the signal matrix for plotting
scaled_signal_H3K9me3 <- t(scale(t(sig_H3K9me3_signal)))  
# Scale rows (regions) - so each region is mean 0, sd 1

pheatmap(scaled_signal_H3K9me3, main = "H3K9me3 (Naive & Exhausted) Significant Regions", show_rownames = FALSE, cutree_rows = 2)

```

Now that we have visualized the changes based on the log2FC we want to have a closer look and plot the 200 peaks with the most significant FC for each Histone Mark for each comparison (part1, part2, part3)

#H3K27ac top 200
```{r}
log2FC_sig_H3K27ac <- log2FC_H3K27ac[sig_rows_H3K27ac]
top200_idx_H3K27ac <- order(abs(log2FC_sig_H3K27ac), decreasing = TRUE)[1:200]
top200_scaled_signal_H3K27ac <- scaled_signal_H3K27ac[top200_idx_H3K27ac, ]
pheatmap(top200_scaled_signal_H3K27ac, main = "H3K27ac (Naive & Exhausted) Top 200 Significant Regions", show_rownames = FALSE, cutree_rows = 2)
```
#H3K27me3 top 200
```{r}
log2FC_sig_H3K27me3 <- log2FC_H3K27me3[sig_rows_H3K27me3]
top200_idx_H3K27me3 <- order(abs(log2FC_sig_H3K27me3), decreasing = TRUE)[1:200]
top200_scaled_signal_H3K27me3 <- scaled_signal_H3K27me3[top200_idx_H3K27me3, ]
pheatmap(top200_scaled_signal_H3K27me3, main = "H3K27me3 (Naive & Exhausted) Top 200 Significant Regions", show_rownames = FALSE, cutree_rows = 2)
```
#H3K9me3
```{r}
log2FC_sig_H3K9me3 <- log2FC_H3K27me3[sig_rows_H3K9me3]
top200_idx_H3K9me3 <- order(abs(log2FC_sig_H3K9me3), decreasing = TRUE)[1:200]
top200_scaled_signal_H3K9me3 <- scaled_signal_H3K27me3[top200_idx_H3K9me3, ]
pheatmap(top200_scaled_signal_H3K9me3, main = "H3K9me3(Naive & Exhausted) Top 200 Significant Regions", show_rownames = FALSE, cutree_rows = 2)
```

#H3K4me3
```{r}
log2FC_sig_H3K4me3 <- log2FC_H3K4me3[sig_rows_H3K4me3]
top200_idx_H3K4me3 <- order(abs(log2FC_sig_H3K4me3), decreasing = TRUE)[1:200]
top200_scaled_signal_H3K4me3 <- scaled_signal_H3K4me3[top200_idx_H3K4me3, ]
pheatmap(top200_scaled_signal_H3K4me3, main = "H3K4me3 (Naive & Exhausted) Top 200 Significant Regions", show_rownames = FALSE, cutree_rows = 2)
```

```{r}
# Step 1: Subset log2FC for significant rows
log2FC_sig_H3K4me3 <- log2FC_H3K4me3[sig_rows_H3K4me3]

# Step 2: Get indices of top 200 by absolute log2FC
top200_idx_H3K4me3 <- order(abs(log2FC_sig_H3K4me3), decreasing = TRUE)[1:200]

# Step 3: Subset the scaled signal matrix
top200_scaled_signal_H3K4me3 <- scaled_signal_H3K4me3[top200_idx_H3K4me3, ]

# Step 4: Extract gene symbols from annotation
# Ensure H3K4me3_annot aligns with original peak order
gene_symbols_H3K4me3 <- H3K4me3_annot$SYMBOL[sig_rows_H3K4me3][top200_idx_H3K4me3]

# Optional: Replace NAs with a placeholder
gene_symbols_H3K4me3[is.na(gene_symbols_H3K4me3)] <- "NA"

# Step 5: Set gene symbols as row names
rownames(top200_scaled_signal_H3K4me3) <- gene_symbols_H3K4me3

# Step 6: Plot heatmap with rownames (gene symbols)
pheatmap(top200_scaled_signal_H3K4me3,
         main = "H3K4me3 (Naive & Exhausted) Top 200 Significant Regions",
         show_rownames = TRUE,
         cutree_rows = 2)
rownames(top200_scaled_signal_H3K4me3) <- make.unique(gene_symbols_H3K4me3)

write.csv(top200_scaled_signal_H3K4me3, "H3K4me3_top200_scaled_with_symbols.csv")
```


