---
title: "Histone Analysis"
author: "Valerie Heckendorn"
date: "2025-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
  library(magick)
  library(R.utils) 
  library(MASS)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(valr)
  library(EnrichedHeatmap)
})
```

# Importing the downloaded data
The histone modifications were shared on GEO Accession Viewer as TXT files.

For the analysis, we split the workload in comparing naive and memory cells (part 1), and naive and exhausted cells (part 2). 



```{r}
head(read.table("peaks/NormRC_H3K27ac_wID.txt", header = TRUE))
```

```{r}
setwd("~/Documents/GitHub/EpiGen_Project_Valerie_Florence/Histone_Modification_EXH/peaks")

# H3K27ac
H3K27ac_df <- read.table("peaks/NormRC_H3K27ac_wID.txt", header = TRUE)

# H3K27me3
H3K27me3_df <- read.table("peaks/NormRC_H3K27me3_wID.txt", header = TRUE)

# H3K4me3
H3K4me3_df <- read.table("peaks/NormRC_H3K4me3_wID.txt", header = TRUE)

# H3K9me3
H3K9me3_df <- read.table("peaks/NormRC_H3K9me3_wID.txt", header = TRUE)

# Converting to GRanges

H3K27ac_gr <- GRanges(seqnames = H3K27ac_df$peakN, ranges = IRanges(start = H3K27ac_df$Start, end = H3K27ac_df$End))
H3K27me3_gr <- GRanges(seqnames = H3K27me3_df$Chr, ranges = IRanges(start = H3K27me3_df$Start, end = H3K27me3_df$End))
H3K4me3_gr <- GRanges(seqnames = H3K4me3_df$Chr, ranges = IRanges(start = H3K4me3_df$Start, end = H3K4me3_df$End))
H3K9me3_gr <- GRanges(seqnames = H3K9me3_df$Chr, ranges = IRanges(start = H3K9me3_df$Start, end = H3K9me3_df$End))

# Create GRanges List

peaks <- GRangesList(
  H3K27ac = H3K27ac_gr,
  H3K27me3 = H3K27me3_gr,
  H3K4me3 = H3K4me3_gr,
  H3K9me3 = H3K9me3_gr
)

```

# We want to visualize changes in histone modifications between the different cells (here: part 2, naive vs. exhausted). The intensity is defined as the mean expression of the histones over the 4 reps. We then want to plot only regions with significant changes, which we defined as log2 fold change >0.1 or <-0.1.

# H3K27ac
```{r}
#mean intensity (all reps) per cell type
H3K27ac_new <- H3K27ac_df[, -c(1:6)]

mean_naive_H3K27ac <- rowMeans(H3K27ac_new[, 1:4])   # naive cols
mean_exh_H3K27ac <- rowMeans(H3K27ac_new[, 13:16])  # memory cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K27ac <- log2((mean_exh_H3K27ac + pseudocount) / (mean_naive_H3K27ac + pseudocount))

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K27ac <- which(abs(log2FC_H3K27ac) > 0.1)

# Subset log2FC vector to significant regions
log2FC_sig <- log2FC_H3K27ac[sig_rows_H3K27ac]

# Convert to 1-column matrix
log2FC_matrix <- matrix(log2FC_sig, ncol = 1)
rownames(log2FC_matrix) <- paste0("peak_", sig_rows_H3K27ac)
colnames(log2FC_matrix) <- "log2FC"

# Plot heatmap with fixed scale from -2 to 2
pheatmap(
  log2FC_matrix,
  main = "H3K27ac log2 Fold Change (Exhausted / Naive)",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE
)
```
# H3K27me3
```{r}
#mean intensity (all reps) per cell type
H3K27me3_new <- H3K27me3_df[, -c(1:6)]

mean_naive_H3K27me3 <- rowMeans(H3K27me3_new[, 1:4])   # naive cols
mean_exh_H3K27me3 <- rowMeans(H3K27me3_new[, 13:16])  # memory cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K27me3 <- log2((mean_exh_H3K27me3 + pseudocount) / (mean_naive_H3K27me3 + pseudocount))

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K27me3 <- which(abs(log2FC_H3K27me3) > 0.1)

# Subset log2FC vector to significant regions
log2FC_sig <- log2FC_H3K27me3[sig_rows_H3K27me3]

# Convert to 1-column matrix
log2FC_matrix <- matrix(log2FC_sig, ncol = 1)
rownames(log2FC_matrix) <- paste0("peak_", sig_rows_H3K27me3)
colnames(log2FC_matrix) <- "log2FC"

# Plot heatmap with fixed scale from -2 to 2
pheatmap(
  log2FC_matrix,
  main = "H3K27me3 log2 Fold Change (Exhausted / Naive)",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE
)
```

#H3K4me
```{r}
#mean intensity (all reps) per cell type
H3K4me3_new <- H3K4me3_df[, -c(1:6)]

mean_naive_H3K4me3 <- rowMeans(H3K4me3_new[, 1:4])   # naive cols
mean_exh_H3K4me3 <- rowMeans(H3K4me3_new[, 13:16])  # memory cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K4me3 <- log2((mean_exh_H3K4me3 + pseudocount) / (mean_naive_H3K4me3 + pseudocount))

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K4me3 <- which(abs(log2FC_H3K4me3) > 0.1)

# Subset log2FC vector to significant regions
log2FC_sig <- log2FC_H3K4me3[sig_rows_H3K4me3]

# Convert to 1-column matrix
log2FC_matrix <- matrix(log2FC_sig, ncol = 1)
rownames(log2FC_matrix) <- paste0("peak_", sig_rows_H3K4me3)
colnames(log2FC_matrix) <- "log2FC"

# Plot heatmap with fixed scale from -2 to 2
pheatmap(
  log2FC_matrix,
  main = "H3K4me3 log2 Fold Change (Exhausted / Naive)",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE
)
```

#H3K9me3
```{r}
#mean intensity (all reps) per cell type
H3K9me3_new <- H3K9me3_df[, -c(1:6)]

mean_naive_H3K9me3 <- rowMeans(H3K9me3_new[, 1:4])   # naive cols
mean_exh_H3K9me3 <- rowMeans(H3K9me3_new[, 13:16])  # memory cols

# Compute log2 fold change (memory / naive), add small pseudocount to avoid log2(0)
pseudocount <- 1e-6
log2FC_H3K9me3 <- log2((mean_exh_H3K9me3 + pseudocount) / (mean_naive_H3K9me3 + pseudocount))

# Filter rows with log2FC > 0.1 (you can add abs(log2FC) > 0.1 if you want both directions)
sig_rows_H3K9me3 <- which(abs(log2FC_H3K9me3) > 0.1)

# Subset log2FC vector to significant regions
log2FC_sig <- log2FC_H3K9me3[sig_rows_H3K9me3]

# Convert to 1-column matrix
log2FC_matrix <- matrix(log2FC_sig, ncol = 1)
rownames(log2FC_matrix) <- paste0("peak_", sig_rows_H3K9me3)
colnames(log2FC_matrix) <- "log2FC"

# Plot heatmap with fixed scale from -2 to 2
pheatmap(
  log2FC_matrix,
  main = "H3K9me3 log2 Fold Change (Exhausted / Naive)",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  breaks = seq(-2, 2, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE
)
```